---
title: "Preservation of UMAP in global structure manifold"
author: "Beatrice, Catherine and Pujan"
date: "2022-11-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
```

# Introduction
Uniform Manifold Approximation and Projection (UMAP) is an algorithm for dimensional reduction.The package that will be developed is suited to process small data with approximately 5,000 data points.

# Loading packages
```{r}
library(umap)
library(tidyverse)
library(magrittr)
```

# Importing the dataset
```{r}
data("iris")
iris
```

# Visualizing the dataset
```{r}
ggplot(data = iris, aes(x = Petal.Length, y = Petal.Width))+
  xlab("Petal Length")+
  ylab("Petal Width") +
  geom_point(aes(color = Species,shape=Species))+
  geom_smooth(method='lm')+
  ggtitle("Iris Petal Length vs Width")
```

# Preprocessing the data for umap
```{r}
# Adding a unique row ID
iris <- iris %>% 
  mutate(ID=row_number())

# Creating a dataframe with all categorical variables with the unique row ID

iris_meta <- iris %>%
  select(ID, Species)

```

# Performing UMAP
```{r}
set.seed(142)
umap_fit <- iris %>%
  select(where(is.numeric)) %>%
  column_to_rownames("ID") %>%
  scale() %>% 
  umap()
umap_df <- umap_fit$layout %>%
  as.data.frame()%>%
  rename(UMAP1="V1",
         UMAP2="V2") %>%
  mutate(ID=row_number())%>%
  inner_join(iris, by="ID")
umap_df
```

# UMAP Plot
```{r}
original_umap_plot <- umap_df %>%
  ggplot(aes(x = UMAP1, 
             y = UMAP2, 
             color = Species))+
  geom_point()+
  labs(x = "UMAP1",
       y = "UMAP2",
      subtitle = "UMAP Plot")
original_umap_plot
```

# Checking Aim 1

To quantify the global structure by establishing preservation of distances between clusters.

## Correlation of original with reconstructed distances between data points

Null hypothesis:- There is no correlation between original and reconstructed distances between data points.

Alternative hypothesis:- There is correlation between original and reconstructed distances between data points.

```{r}
# Pearson is most appropriate for measurements taken from an interval scale, while the Spearman is more appropriate for measurements taken from ordinal scales.
cor.test(iris$Petal.Length, umap_df$UMAP1, method = 'pearson')
cor.test(iris$Petal.Width, umap_df$UMAP2, method = 'pearson')

```

The result shows the p-value of less than 0.05 hence we reject the null hypothesis and conclude that there is a correlation between original and reconstructed distances between data points.

## Preservation of pairwise Mahalanobis distances between clusters
```{r}
g1 <- iris[3:4]
g2 <- umap_df[1:2]
D.sq <- function (g1, g2) {
    dbar <- as.vector(colMeans(g1) - colMeans(g2))
    S1 <- cov(g1)
    S2 <- cov(g2)
    n1 <- nrow(g1)
    n2 <- nrow(g2)
    V <- as.matrix((1/(n1 + n2 - 2)) * (((n1 - 1) * S1) + ((n2 - 1) * S2)))
    D.sq <- t(dbar) %*% solve(V) %*% dbar
    res <- list()
    res$D.sq <- D.sq
    res$V <- V
    res
}

D.sq(g1,g2)
```

The mahalanobis distance between clusters is 3.7158. We will therefore aim to achieve mahalanobis Distance of 1 or lower so that the data point are right among the benchmark points.




