---
title: "umapgp-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{umapgp-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(UMAPgp)
```
## Package Vignette: Rationale
UMAP preserves local structure by grouping neighboring data points together which provides an informative visualization of heterogeneity in a dataset. However, if one wants to understand the relationship between clusters of data points, this can be troublesome since UMAP does not warrant that inter-cluster distances are preserved correctly. To address the shortcomings, we have created a function, UMAPgp, and developed its equivalent package which when applied preserves the global structure of manifold learning algorithms for small datasets. It also preservs the global structure of manifold learning algorithms without losing the finer details.The package that will be developed is suited to process small data with approximately 1,000 data points.

## Package Vignette: Significance/ Innovation
Preservation of global structure in manifold learning algorithms will ensure that UMAP represents the big-picture while not losing the fine details hence, better visualization. Based on the project results, we will modify the cost function shown below and create an R package of a final embedding with an optimal cost function.

## Addressing Aim 1 and Aim 2: Preservation of Distances between the Clusters/Species

In this section, we shall load the required packages and import the dataset 'Iris' which shall be used for the purpose of this project. We will work towards visualizing the dataset alongwith preprocessing it to perform UMAP. Additionally, we shall create the UMAP plot and check if the distances between the clusters/species have been preserved. It shall address our Aim 1 and Aim 2. 

```{r}
# Loading the packages
library(umap)
library(tidyverse)
library(magrittr)

# Importing the dataset
data("iris")
iris

# Visualizing the dataset
ggplot(data = iris, aes(x = Petal.Length, y = Petal.Width))+
  xlab("Petal Length")+
  ylab("Petal Width") +
  geom_point(aes(color = Species,shape=Species))+
  geom_smooth(method='lm')+
  ggtitle("Iris Petal Length vs Width")

# Pre-processing the data for umap by adding a unique row ID and creating a dataframe with all categorical variables with the unique row ID

iris <- iris %>% 
  mutate(ID=row_number())

iris_meta <- iris %>%
  select(ID, Species)

# Performing UMAP
set.seed(142)
umap_fit <- iris %>%
  select(where(is.numeric)) %>%
  column_to_rownames("ID") %>%
  scale() %>% 
  umap()
umap_df <- umap_fit$layout %>%
  as.data.frame()%>%
  rename(UMAP1="V1",
         UMAP2="V2") %>%
  mutate(ID=row_number())%>%
  inner_join(iris_meta, by="ID")
umap_df

# UMAP Plot
umap_df %>%
  ggplot(aes(x = UMAP1, 
             y = UMAP2, 
             color = Species))+
  geom_point()+
  labs(x = "UMAP1",
       y = "UMAP2",
      subtitle = "UMAP plot")

